To be FAIR: Theory Specification Needs an Update

### 2025-02-21 manuscript.Rmd
closes #1

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and reusable [REF] for both humans and computers.
Since their inception in 2014, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.

@@

The so-called "replication crisis" has prompted extensive reforms in social science [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.

@@

Scholars have been raising concerns about the state of theory in social science for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that social scientific theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories do not make precise predictions,
they are hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

@@

The present paper introduces open science methods for theory specification.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts compliant with the FAIR principles.
The resulting FAIR theories are archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo). 
FAIR theories are *Findable* via a DOI or search,
*Accessible* in a machine- and human-readable filetype,
*Interoperable* within the data analysis environment,
and *Reusable* in the practical and legal sense, so that they may be iteratively improved - potentially via a participative process.
Digital representations of theory intentionally is a broad term, particularly including textual representations of a given theory, as well as formal representations, such as mathematical notation, algorithmic pseudo code, or a set of logical clauses. <!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in the social sciences.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [REF Fried].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
-<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->

@@

Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
the method is not contained to any one particular definition.
It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, the nature of the information tracked will likely change.
For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
From this theory, more precise *statistical/mathematical models* could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Then, an even more specific *generative/computational model* could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g. longitudinal designs observing change over time), measurement tools (e.g. different questionnaires used to assess the same construct), or study subjects (e.g. specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

Although we are in favor of the advancement towards formal theories, 
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).

@@

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
At present, the primary unit of social scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the [DataCite Metadata Schema]() offers a standard way to document the nature of relationships between information artifacts.
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`. Other types are useful for making relationships between multiple theory objects explicit: If an existing narrative theory is FAIRified without much own contribution, the FAIR theory would cross-reference the paper with the narrative theory as `relationType: IsDerivedFrom`; if substantial own contributions were necessary, `relationType:IsNewVersionOf` could be used. If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->

@@

The replication crisis has brought the inadequacies of contemporary theoretical practices in the social sciences into focus.
Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited Interoperability, and are rarely Reused.
These limitations impede cumulative knowledge production in our field,
leading to an accumulation of "one-shot" empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
We demonstrated how to create, version-control, and archive theories as digital information artifacts.
We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.

Making theory FAIR allows researchers to more easily find a relevant framework;
access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
Whereas the idea of theory can be quite nebulous to empirical social scientists,
FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
While FAIR theory does not directly reduce ambiguity,
it provides a framework within which scholars can iteratively increase precision and formalization.
The FAIR principles also facilitate new ways of collaboration,
leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
<!-- These practices align with the open science ethos, ensuring theories evolve dynamically based on empirical evidence and collaborative input. -->

@@

A limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [@frankenhuisStrategicAmbiguitySocial2023] and lack of theory formalization [@robinaughInvisibleHandsFine2021].
However, our work does establish a framework that allows for and promotes the formalization of theories. 
The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
Another limitation of scope is that FAIR theory does not resolve other related issues in social sciences, such as the measurement crisis [@bringmannBackBasicsImportance2022] and lack of standardized ontologies for psychological constructs [@boscoMetaBUSVehicleFacilitating2017].
However, our work here provides a template for addressing such challenges, <!-- NvD: I am not sure if we are offering a template to address these challenges. If we do, we should make this explicit. -->
and any advancements in the areas of measurement and ontology will serve to amplify the value of FAIR theories, particularly when such resources are cross-referenced in the metadata (e.g., on Zenodo).

@@

One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.

@@

FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
It provides much-needed open science methods for the inductive phase of the empirical cycle,
closing a critical gap in the scientific process.
FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
This paves the way for more theory-driven scholarship,
and accelerates cumulative knowledge acquisition in the social sciences and beyond.

>>>>

The FAIR Guiding Principles (hereafter: FAIR principles) were established to improve the reusability of research data by making them more Findable, Accessible, Interoperable and Reusable [@wilkinsonFAIRGuidingPrinciples2016] for both humans and computers.
Since the FAIR principles' inception, scholars have demonstrated their relevance for making other information artefacts more open, such as research software [@lamprechtFAIRPrinciplesResearch2019] and computational workflows [@wilkinson2024applying].
This paper argues that the FAIR principles can advance effective and transparent scholarly communication about theory.
To this end, we introduce "FAIR theory":
a digital instantiation of a scientific theory, published as a self-contained and citable information artifact distinct from the scientific paper,
compliant with the FAIR principles.
FAIR theory has the potential to improve the efficiency of scholarly communication and
accelerate cumulative knowledge acquisition.
We focus on applications in psychology, but the principles are relevant across the social sciences, and beyond.

@@

The so-called "replication crisis" has prompted extensive reforms in psychology and other scientific fields [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
Under the original definition, few would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience
Yet, under the expanded framework, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.

@@

Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack precision, or  *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories do not make precise predictions,
they are hard to falsify and difficult to understand on their own,
without either substantial interpretation or additional background knowledge.
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with apropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
FAIR theories are *Findable* via a DOI or search,
*Accessible* in a machine- and human-readable filetype,
*Interoperable* within the data analysis environment,
and *Reusable* in the practical and legal sense, so that they may be iteratively improved - potentially via a participative process.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see Table \@ref(tab:tabfair).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

The semantic view challenges the necessity of distinct domain languages for different scientific fields, and instead advocates for formalizing theories using mathematics.
It shifts the focus from theories as collections of sentences to mathematical models.
The term "model" is not uniquely defined within the literature;
it has been described as a "specific instantiation of theory narrower in scope and often more concrete, commonly applied to a particular aspect of a given theory" [@friedTheoriesModelsWhat2020].
This implies that theories and models are not fundamentally distinct, but rather, that for each model, there is a more general theory that subsumes it (one person's model is another person's theory).
-<!-- Paradigms at the broadest level encompass the larger narrative, frameworks gather and organize concepts and terms, but only theories have a degree of specificity that allows us to derive specific testable and falsifiable hypotheses or models with exact mathematical relations between concepts (cf. Partelow, 2023, Table 1, https://link.springer.com/article/10.1007/s13412-023-00833-w#). -->

@@

Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
the method is not contained to any one particular definition.

It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory. 
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, the nature of the information tracked will likely change.
For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
From this theory, more precise *statistical/mathematical models* could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Then, an even more specific *generative/computational model* could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g. longitudinal designs observing change over time), measurement tools (e.g. different questionnaires used to assess the same construct), or study subjects (e.g. specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use. 
<!-- NvD: Maybe we should add "specification" as part of "formalization" or "formalisation" as a part of "specification", because not everybody that is advocating better theories thinks that we should focus on formalization.  -->
Greater formalization increases theories' *empirical content* [@popperLogicScientificDiscovery2002] because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence [@lewandowsky2010computational].
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Compared to theories expressed in natural language,
formal theories facilitate inconsistency checking and evaluation of a theory's (lack of) ambiguity.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal,
and formal theory can be represented in a way that is not FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to in a journal article without any key words to identify it as a theory paper (lacking Findability), represented as a bitmap image (limiting Accessibility and Interoperability), and behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see *Accessibility*).

@@

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`. Other types are useful for making relationships between multiple theory objects explicit: If an existing narrative theory is FAIRified without much own contribution, the FAIR theory would cross-reference the paper with the narrative theory as `relationType: IsDerivedFrom`; if substantial own contributions were necessary, `relationType:IsNewVersionOf` could be used. If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->

@@

The replication crisis has brought the inadequacies of contemporary theoretical practices in psychology and other fields into focus.
Psychological theories often fall short of all FAIR principles: they are hard to find and access, have limited Interoperability, and are rarely Reused.
These limitations impede cumulative knowledge production in our field,
leading to an accumulation of "one-shot" empirical findings, without commensurate advancement in our principled understanding of psychological phenomena.
We argued that applying the FAIR principles to theory offers a structured solution to these shortcomings.
We demonstrated how to create, version-control, and archive theories as digital information artifacts.
We introduced the `theorytools` R-package to partly automate these processes, reducing the barrier of entry for researchers,
and creating a FAIR resource for theory construction tools and documentation that can be continuously updated as best practices develop further.

Making theory FAIR allows researchers to more easily find a relevant framework;
access and understand it; interact with it in a very practical manner, for example, by deriving predictions from it, or using it to select control variables; and reuse it, contributing changes to existing theories or splitting of in a new direction.
Whereas the idea of theory can be quite nebulous to empirical psychologists,
FAIR theory makes theoretical work practical and tangible, incorporating theory into scholars' workflows.
Having a concrete object to iterate upon facilitates the systematic improvement and iterative refinement of psychological theories, thus substantially increasing the efficiency of research.
While FAIR theory does not directly reduce ambiguity,
it provides a framework within which scholars can iteratively increase precision and formalization.
The FAIR principles also facilitate new ways of collaboration,
leveraging tools like Git for version control and Zenodo for archiving to document provenance and facilitate contributions from diverse researchers.
<!-- These practices align with the open science ethos, ensuring theories evolve dynamically based on empirical evidence and collaborative input. -->

@@

A limitation of scope is that FAIR theory does not directly resolve problems related to strategic ambiguity [@frankenhuisStrategicAmbiguitySocial2023] and lack of theory formalization [@robinaughInvisibleHandsFine2021].
However, our work does establish a framework that allows for and promotes the formalization of theories. 
The example of the empirical cycle demonstrates how FAIR principles can guide theory formalization and foster cumulative progress.
Another limitation of scope is that FAIR theory does not resolve other related issues in psychology, such as the measurement crisis [@bringmannBackBasicsImportance2022] and lack of standardized ontologies for psychological constructs [@boscoMetaBUSVehicleFacilitating2017].
However, our work here provides a template for addressing such challenges, <!-- NvD: I am not sure if we are offering a template to address these challenges. If we do, we should make this explicit. -->
and any advancements in the areas of measurement and ontology will serve to amplify the value of FAIR theories, particularly when such resources are cross-referenced in the metadata (e.g., on Zenodo).

@@

One issue that intersects with FAIR theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [@bringmannBackBasicsImportance2022].
Additionally, the "jingle-jangle" fallacy is prevalent in psychology:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of measurement.

@@

FAIR theory is a major step forward towards more transparent, collaborative, and efficient theory construction.
It provides much-needed open science methods for the inductive phase of the empirical cycle,
closing a critical gap in the scientific process.
FAIR theory makes theory more tangible, enabling scholars to incorporate it in their day-to-day workflows in order to derive hypotheses, select control variables, and contribute new data-driven insights.
This paves the way for more theory-driven scholarship,
and accelerates cumulative knowledge acquisition in the psychology, other social sciences, and beyond.
